Finally Tagless Staged Interpreters for Simpler Typed Languages

### 2007-04-06 tagless/tagless-final.tex
lots of little edits here and there, for style and length

Our contributions are as follows.
\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles
   (\S\ref{discussion}).
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation (\S\ref{language}) to partial evaluation and CPS transformation (\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state (\S\ref{state}).
\item We describe an approach to self\hyp interpretation compatible with the
  above (\S\ref{selfinterp}).  Self\hyp interpretation turned out harder than expected.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
state that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We further discuss related work in~\S\ref{related}.

@@

\section{Variations and discussion}\label{discussion}

We show that our |Symantics| and generally our approach accomodates
a number of variants without any (new) difficulties.  In particular,
we show how a call-by-name CPS interpreter, a call-by-value CPS
program transformation, as well as how to deal with imperative features.
Finally, we comment on the requirements on the type system to support 
our approach.

@@

The object language generally inherits the evaluation strategy from
the metalanguage. Evaluating an application |e1 e2| in OCaml requires
the evaluation of |e2| before performing the application. Therefore,
when evaluating the encoding of the object language application 
|app e1 e2|, |e2| will always have to be evaluated. One may wonder if it is
possible to represent a call-by-name (CBN) object language in a
call-by-value (CBV) metalanguage. The answer, given already by 
\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is yes thanks to CPS, which they
specifically introduced 
to make the evaluation strategy of a definitional interpreter
independent of the strategy of the metalanguage. We demonstrate the
same independence in the typed setting. Specifically, we
demonstrate CBN evaluation for our object language in OCaml, which is
call-by-value. We do so by building a CBN CPS interpreter for the object
language.

The CBN CPS interpretation was introduced by \citet{PlotkinCBN}. As in
CBV CPS, the interpretation of an object term is a function
mapping a continuation~|k| to the answer
returned by~|k|.
\begin{code}
let int (x:int) = fun k -> k x
let add e1 e2 = fun k ->
  e1 (fun v1 -> e2 (fun v2 -> k (v1 + v2)))
\end{code}
In both cases, the returned value has type 
|(int -> 'w)->'w| where |'w| is the (polymorphic) answer type.

The interpretation of abstraction and application in CBN CPS is
different from CBV CPS and is as follows:
\begin{code}
let lam f = fun k -> k f
let app e1 e2 = fun k -> e1 (fun f -> (f e2) k)
\end{code}
In the tell-tale sign of CBN, in interpreting the application
we do not `evaluate' the argument e2, that is, we do apply it to the
continuation. Rather, we pass the unevaluated |e2| to the abstraction.
The result of |lam (fun x -> add x (int 1))| has the type
|((((int->'w1)->'w1) -> ((int->'w1)->'w1)) -> 'w2) -> 'w2|.
We want to collect the above interpretation functions into a module
with the |Symantics| signature, so as to fit the CBN CPS interpreter within our
general framework. Alas, we encounter the same problem we had in
Section~\ref{S:PE-problem}, in writing a type equation for the |repr|
type that can represent both the type of |int i| and the type of 
|lam f|. We cannot say that the result of interpreting an object
expression of type |t| is |(t->'w)->'w| because if |t| happens to
be an arrow type (such as |int->int| in the above example), the result
type obviously does not fit the desired pattern. We need a type
function with a typecase distinction. In Section~\ref{S:PE-solution} we have
shown how to encode such a function, using the extra type
argument |'sv| of the |repr| type. Furthermore, we notice that the type 
function |repr_pe| needed for the partial evaluator 
(in~\S\ref{S:PE-solution}) is precisely the same type function we
need for CBN CPS. This gives us for the following CBN CPS interpreter,
with the |Symantics| signature:

@@

We have made our interpreter fully polymorphic over the answer type.
We could have just as easily made |RCN| a functor, parameterized over
the answer type -- as is common when writing CPS in SML. In the
present case, we chose to use the higher-rank polymorphism of offered by
OCaml record types.

@@

The transformation is canonical, as the following (abbreviated) code
shows
\begin{code}
module CPST(S: Symantics) = struct
  let int i = S.lam (fun k -> S.app k (S.int i))
  let add e1 e2 = S.lam (fun k ->
    S.app e1 (S.lam (fun v1 ->
    S.app e2 (S.lam (fun v2 -> S.app k (S.add v1 v2))))))
  let if_ ec et ef = S.lam (fun k ->
    S.app ec (S.lam (fun vc ->
    S.if_ vc (fun () -> S.app (et ()) k) (fun () -> S.app (ef ())
    k))))
  ...
  let lam f = S.lam (fun k -> S.app k (S.lam (fun x ->
    f (S.lam (fun k -> S.app k x)))))
  let app e1 e2 = S.lam (fun k -> 
    S.app e1 (S.lam (fun f ->
    S.app e2 (S.lam (fun v -> S.app (S.app f v) k)))))
  let fix f  = S.fix (fun self -> (f self))
end
\end{code}
The code literally establishes a mapping from CPS interpretations to
(direct) interpretations, the latter performed by 
the base interpreter |S|.

A reader may notice that the above module does not define the type
equation for |repr| and so it does not have the signature Symantics.
The reason for that is again the result type of |lam f|. Whereas
|int i| or |add e1 e2| have the (abbreviated) 
type |('c,..., (int -> 'w) -> 'w) S.repr|,
the result of |lam (fun x -> add x (int 1))| has type
|('c,...((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr|. 
Hence, to write the type equation for |CPST.repr| we need a 
type-discriminating
function similar to the type function |repr_pe| of
\S\ref{S:PE-solution}. Alas, the function we need here is
similar but not identical to |repr_pe|. Thus we need to add 
another type variable to the |repr| type and modify the Symantics
signature accodringly. Again, the terms in previous Symantics module
stay unchanged, but the |repr| type equations in those structures have to
account for a new (phantom) type variable.

To save space, we show a simplified and less principled approach that
lets us use the module |CPST| as it is. Because the module does not
have the signature Symantics, we cannot apply the |EX| functor to it.
But we can write the tests nevertheless:
\begin{code}
module T = struct
 module M = CPST(C)
 open M
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = ... (* the same as before *)
 let testpowfix7 = (* the same as before *)
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
That is, we instantiate |CPST| with the desired base intrepreter (the
compiler |C| in the case above) so that we can use the result |M| to
interpret object level terms. The sample terms are \emph{exactly}
as we used before. We just copied them over. That copying step is the
price we pay for the simplified treatment. We shall also encounter this
copying when discussing the self-interpreter in the next section. We
will see then that the copying step is not frivolous, but represents
a profound instance of 
``plugging the hole'', which is one of the many faces of polymorphism.

It is instructive to evaluate a few examples. With 
|CPST| instantiated with the compiler~|C| above, our
running test |T.test1 ()| gives
\begin{code}
.<fun x_5 -> ((fun x_2 -> 
        (x_2 (fun x_3 -> fun x_4 -> (x_4 x_3))))
        (fun x_6 -> ((fun x_1 -> (x_1 (true))) 
        (fun x_7 -> ((x_6 x_7) x_5)))))>.
\end{code}

which shows the naive CPS transformation of the original code. The result is
quite unoptimal with several apparent $\beta$-redexes. 
By instantiating |CPST| with the |P| evaluator instead, our
running test gives the expected
\begin{code}
  P.repr = {P.st = Some <fun>;
            P.dy = .<fun x_5 -> (x_5 (true))>.}
\end{code}

@@

Intuitively, self\hyp interpretation is straightforward in our
framework: the functions comprising the interpreters
in~\S\ref{S:interpreter-RL} may just as well be written in our object
language.  In particular, the following \emph{object} functions implement an
evaluator.  (We use the number~$0$ in lieu of a unit value.)
\begin{align*}
    \ident{int} &= \fun{x} x &
    \ident{lam} &= \fun{f} f \\
    \ident{add} &= \fun{x} \fun{y} x+y &
    \ident{app} &= \fun{f} \fun{x} fx \\
    \ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0} &
    \ident{fix} &= \fun{g} \fix{f} \fun{x} gfx
\end{align*}
We thus map each object terms~$e$ to an object term~$\encode{e}$ as follows.
We call this mapping \emph{pre-encoding}.
\begin{align*}
    \encode{x} &= x &
    \encode{\fun{x}e} &= \ident{lam} (\fun{x} \encode{e}) \\
    \encode{n} &= \ident{int}\, n &
    \encode{fx} &= \ident{app} \encode{f} \encode{x} \\
    \encode{e_1 + e_2} &= \ident{add} \encode{e_1} \encode{e_2} &
    \encode{\fix{f}e} &= \ident{fix} (\fun{f} \encode{e}) \\
    \encode{\cond{b}{t}{e}} &= \rlap{$\ident{if\_} \encode{b}
        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$}
\end{align*}
The metavariables $x$ and~$n$ stand for a variable and an integer,
respectively.
This pre-encoding is just like how we represent object terms in the
metalanguage in the preceding sections, but it produces
terms in the object language rather than metalanguage.

To evaluate~$\encode{e}$, then, we
instantiate the free variables in~$\encode{e}$ such as $\ident{int}$,
$\ident{lam}$, and $\ident{add}$ by their definitions above.  For
example, the familiar object term $(\fun{x}x)\True$ pre-encodes to
\begin{equation*}
    \encode{(\fun{x}x)\True} = \ident{app}
    (\ident{lam} (\fun{x} x))\, (\ident{bool} \True),
\end{equation*}
and to evaluate this pre-encoded term is to evaluate the object term
\begin{equation*}
    (\fun{f} \fun{x} fx) \,
    ((\fun{f} f) (\fun{x} x))\, ((\fun{b} b) \True).
\end{equation*}
Because the evaluator above mostly consists of glorified identity
functions, we expect even our simple partial evaluator to reduce the
result of this instantiation to~$e$.  In general, to
interpret~$\encode{e}$ using an interpreter is to instantiate these free
variables by that interpreter's definitions.

@@

Any approach to self\hyp interpretation needs to spell out first how to
encode object terms~$e$ to object terms~$\Encode{e}$, and then how to
interpret~$\Encode{e}$ in the object language.  For our approach, we
want to define encoding in terms of pre-encoding, and interpretation
using some notion of instantiation.  Unfortunately, the simple type
structure of our object language hinders both tasks.  To continue with
the example term above, we could try to define
\begin{equation*}
    \Encode{e} =
    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}} \encode{e},
\end{equation*}
in particular
\begin{equation*}
    \Encode{(\fun{x}x)\True} =
    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool} \True).
\end{equation*}
To type-check this encoded term, we give the variable $\ident{lam}$ the
simple type $(\BB\to\BB)\to\BB\to\BB$.
We then define the self\hyp interpreter
\begin{equation*}
    \si = \fun{e} e
    (\fun{f} \fun{x} fx)
    (\fun{f} f)
    (\fun{b} b)
\end{equation*}
and apply it to the encoded term.  The result is the object term
\begin{multline*}
    \bigl(\fun{e} e (\fun{f} \fun{x} fx) (\fun{f} f) (\fun{b} b)\bigr)
\\
    \bigl(
    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool} \True)
    \bigr),
\end{multline*}
which partially evaluates to~$\True$ easily.
However, encoding fails on an term with multiple $\lambda$\hyp
abstractions at different types.  For example, the pre-encoding
\begin{equation*}
    \encode{\fun{f}\fun{x}fx}
    = \ident{lam} (\fun{f} \ident{lam} (\fun{x} \ident{app} f x))
\end{equation*}
does not type-check in any typing environment, because $\ident{lam}$ needs
to take two incompatible types.  In sum, we need polymorphism in the
object type system to type $\ident{lam}$, $\ident{app}$, $\ident{fix}$,
and~$\ident{if\_}$ (and $\Encode{e}$ and~$\si$).
(The polytypes in |Symantics| are where Haskell type classes and OCaml
modules supply this polymorphism.)  Moreover, we need to encode any
polymorphism in the object language into the object language to achieve
self\hyp interpretation.

@@

Instead of adding higher-rank and higher-kind polymorphism to our object language (along
with polymorphism over kinds), we add let-bound polymorphism.  As usual,
we can add a new typing rule
\begin{equation*}
    \begin{prooftree}
        e_1:t_1 \quad \subst{e_2}{x}{e_1}:t_2
        \justifies \be{x=e_1} e_2 : t_2
    \end{prooftree}
    .
\end{equation*}
The pre-encoding of a let\hyp expression is trivial.
\begin{align*}
    \encode{\be{x=e_1}e_2} \quad &= \quad \be{x=\encode{e_1}} \encode{e_2}
\intertext{A \emph{context} is an object term with a hole~$[~]$.  The
hole may occur under a binder, so plugging a term into the context may
capture free variables of the term.  By pre-encoding a hole to a hole,
we extend pre-encoding from a translation on terms to one on
contexts.}
    \encode{[~]} \quad &= \quad [~]
\end{align*}
We define an interpreter in the object language to be not a term but
a context.  For example, the evaluator is the context
\begin{align*}
    \si[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} x&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y&} \\
        \be{\ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}&} \\
        \be{\ident{lam} &= \fun{f} f&} \\
        \be{\ident{app} &= \fun{f} \fun{x} fx&} \\
        \be{\ident{fix} &= \fun{g} \fix{f} \fun{x} gfx&} [~],
    \end{tabular}
\intertext{and the size\hyp measurer is the context}
    \mathrm{SZ}[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} 1&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y+1&} \\
        \be{\ident{if\_}&= b + t\,0 + e\,0&} \\
        \be{\ident{lam} &= \fun{f} f\,0 + 1&} \\
        \be{\ident{app} &= \fun{f} \fun{x} f + x + 1&} \\
        \be{\ident{fix} &= \fun{g} g\,0 + 1&} [~].
    \end{tabular}
\end{align*}
To interpret an object term~$e$ using an interpreter $I[~]$ is to
evaluate the object term~$I[\encode{e}]$.  $\si$ is
a self\hyp interpreter in the following sense.
\begin{prop}
    $\si[\encode{e}]$ is observationally equivalent to~$e$.
\end{prop}
As a corollary, we can pre-encode the self\hyp interpreter itself as
a context: the term $\si[\encode{\si[\encode{e}]}]$ is observationally
equivalent to $\si[\encode{e}]$, and in turn to~$e$.  In other words,
$\si$ can interpret itself.

Our partial evaluator is optimal with respect to the self\hyp
interpreter~$\si$.
\begin{prop}
    Let $\pe$ be the partial evaluator~|P| in~\S\ref{S:PE-solution}.
    Then the object terms $\pe(\si[\encode{e}])$ and~$\pe(e)$ are
    either both undefined or both defined and
    equal up to $\alpha$\hyp conversion.
%    \ccshan{Check this}
%    \jacques{seems correct.  If it said that the results are operationally
%    equivalent to e itself, then that would be less clear.}
\end{prop}

@@

We type-check a pre-encoded term~$\encode{e}$ and its interpreter~$I[~]$
only together, never separately.  This treatment has the drawback that
we must duplicate a pre-encoded term in order to interpret it in
multiple ways.  The meta-notions of contexts and plugging may seem ad
hoc, but in fact they are explications of the type-class and module
machinery that we have been using in this paper all along.

@@

\citet{Danvy-tagging-encoding} discusses Jones optimality at length.
There, the authors demonstrate the usefulness of HOAS for typed
self-interpretation.  However, no claim is made that the source program
is itself typed -- just that the interpreter is.  Our reading is that
the source language is not typed.  However, in sharp contrast with 
their interpreter, none of our evaluators can product pattern-match errors
(unlike the code that accompanies \citet{Danvy-tagging-encoding}).
In other words, their language (and interpreter) does use tags, however
the shift to HOAS allows the partial evaluator to remove all the tags.
In contrast, we have no tags at all.

>>>>

Our contributions are as follows.
\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles
   (\S\ref{variations}).
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation (\S\ref{language}) to partial evaluation and CPS transformation (\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state (\S\ref{state}).
\item We describe an approach to self\hyp interpretation compatible with the
  above (\S\ref{selfinterp}).  Self\hyp interpretation turned out harder than expected.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
state that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We further discuss related work in~\S\ref{related}.

@@

\section{Variations}\label{variations}

We show that our |Symantics| and generally our approach accomodates
a number of variants without any (new) difficulties.  In particular,
we show how a call-by-name CPS interpreter, a call-by-value CPS
program transformation.  Finally, we comment on the requirements on the type
system to support our approach.

@@

The object language generally inherits the evaluation strategy from
the metalanguage -- call-by-value (CBV) in OCaml, call-by-name (CBN) in Haskell.
One may wonder if it is
possible to represent a call-by-name object language in a
call-by-value metalanguage. The answer, given already by 
\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is yes
thanks to CPS, which they specifically introduced 
to make the evaluation strategy of a definitional interpreter
independent of the strategy of the metalanguage. We demonstrate the
same independence in the typed setting. Specifically, we
demonstrate CBN evaluation for our object language in OCaml.
We do so by building a CBN CPS interpreter for the object
language.

The interpretation of an object term is a function
mapping a continuation~|k| to the answer
returned by~|k|.
\begin{code}
let int (x:int) = fun k -> k x
let add e1 e2 = fun k ->
  e1 (fun v1 -> e2 (fun v2 -> k (v1 + v2)))
\end{code}
In both cases, the returned value has type 
|(int -> 'w)->'w| where |'w| is the (polymorphic) answer type.

The interpretation of abstraction and application in CBN CPS is
different from CBV CPS and is as follows:
\begin{code}
let lam f = fun k -> k f
let app e1 e2 = fun k -> e1 (fun f -> (f e2) k)
\end{code}
Charateristic of CBN, interpreting the application does not
``evaluate'' the argument |e2|, that is, we do apply it to the
continuation. Rather, we pass the unevaluated |e2| to the abstraction.
The result of |lam (fun x -> add x (int 1))| has the type
|((((int->'w1)->'w1) -> ((int->'w1)->'w1)) ->'w2)->'w2|.
We would like to collect those interpretation functions into a module
with signature |Symantics|, to include the CBN CPS interpreter within our
general framework. Alas, we encounter the same problem we had in
\S\ref{S:PE-problem}.
We cannot say that the result of interpreting an object
expression of type |t| is |(t->'w)->'w| because if |t| happens to
be an arrow type, the result
type obviously does not fit the desired pattern. We need a type
function with a typecase distinction.  Our previous solution 
(\S\ref{S:PE-solution}) where we use an extra type
argument |'sv| of the |repr| type, to encode such type-level functions,
also works here.  Furthermore, we notice that the type 
function |repr_pe| needed for the partial evaluator 
(in~\S\ref{S:PE-solution}) is precisely the same type function we
need for CBN CPS.

@@

We have made our interpreter fully polymorphic over the answer type.
We could have also made |RCN| a functor, parameterized over
the answer type -- as is common when writing CPS in SML. Here
we chose to use the higher-rank polymorphism offered by
OCaml record types.

@@

The transformation is canonical, as the following (abbreviated) code
shows
\begin{code}
module CPST(S: Symantics) = struct
  let int i = S.lam (fun k -> S.app k (S.int i))
  let add e1 e2 = S.lam (fun k ->
    S.app e1 (S.lam (fun v1 ->
    S.app e2 (S.lam (fun v2 -> S.app k (S.add v1 v2))))))
  ...
  let lam f = S.lam (fun k -> S.app k (S.lam (fun x ->
    f (S.lam (fun k -> S.app k x)))))
  let app e1 e2 = S.lam (fun k -> 
    S.app e1 (S.lam (fun f ->
    S.app e2 (S.lam (fun v -> S.app (S.app f v) k)))))
  let fix f  = S.fix (fun self -> (f self))
end
\end{code}
The code explicitly establishes a mapping from CPS interpretations to
(direct) interpretations, as performed by 
the base interpreter |S|.

A reader may notice that the above module does not define |repr|
and thus does not have signature |Symantics|.
The reason for that is again the result type of |lam f|. Whereas
|int i| or |add e1 e2| have the (abbreviated) 
type |('c,..., (int -> 'w) -> 'w) S.repr|,
the result of |lam (fun x -> add x (int 1))| has type\\
|('c,...((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr|. 
Hence, to write the type equation for |CPST.repr| we again need a 
type-discriminating
function similar to the type function |repr_pe| of
\S\ref{S:PE-solution}. Alas, the function we need here is
similar but not identical to |repr_pe|. Thus we need to add 
another type variable to the |repr| type and modify the Symantics
signature accodringly. Again, the terms in previous Symantics module
stay unchanged, but the |repr| type equations in those structures have to
account for a new (phantom) type variable.

To save space, we show a simplified and less principled approach that
lets us use the module |CPST| as is. Because the module does not
have the signature |Symantics|, we cannot apply the |EX| functor to it.
But we can nevertheless write the tests:
\begin{code}
module T = struct
 module M = CPST(C)
 open M
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = ... (* the same as before *)
 let testpowfix7 = (* the same as before *)
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
That is, we instantiate |CPST| with the desired base intrepreter (the
compiler |C| in the case above) so that we can use the result |M| to
interpret object level terms. Those terms are \emph{exactly} as before.
Having to textually copy the terms is the
price we pay for this simplified treatment. We shall also encounter this
copying when discussing self-interpretation.  We
will see then that the copying step is not frivolous, but represents
a profound instance of 
``plugging the hole'', which is one of the many faces of polymorphism.

It is instructive to evaluate a few examples. With 
|CPST| instantiated with the compiler~|C| above,
|T.test1| gives
\begin{code}
.<fun x_5 -> ((fun x_2 -> 
        (x_2 (fun x_3 -> fun x_4 -> (x_4 x_3))))
        (fun x_6 -> ((fun x_1 -> (x_1 (true))) 
        (fun x_7 -> ((x_6 x_7) x_5)))))>.
\end{code}

\noindent which shows the na\"{\i}ve CPS transformation of the original code.
The result is sub-optimal with several apparent $\beta$-redexes. 
By instantiating |CPST| with |P| instead, we get the expected
\begin{code}
  P.repr = {P.st = Some <fun>;
            P.dy = .<fun x_5 -> (x_5 (true))>.}
\end{code}

@@

Intuitively, self\hyp interpretation is straightforward in our
framework: the functions comprising the interpreters
in~\S\ref{S:interpreter-RL} may just as well be written in our object
language.  In particular, the following \emph{object} functions implement an
evaluator.  (We use the number~$0$ in lieu of a unit value.)
\begin{align*}
    \ident{int} &= \fun{x} x &
    \ident{lam} &= \fun{f} f \\
    \ident{add} &= \fun{x} \fun{y} x+y &
    \ident{app} &= \fun{f} \fun{x} fx \\
    \ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0} &
    \ident{fix} &= \fun{g} \fix{f} \fun{x} gfx
\end{align*}
We thus map each object terms~$e$ to an object term~$\encode{e}$ as follows.
We call this mapping \emph{pre-encoding}.
\begin{align*}
    \encode{x} &= x &
    \encode{\fun{x}e} &= \ident{lam} (\fun{x} \encode{e}) \\
    \encode{n} &= \ident{int}\, n &
    \encode{fx} &= \ident{app} \encode{f} \encode{x} \\
    \encode{e_1 + e_2} &= \ident{add} \encode{e_1} \encode{e_2} &
    \encode{\fix{f}e} &= \ident{fix} (\fun{f} \encode{e}) \\
    \encode{\cond{b}{t}{e}} &= \rlap{$\ident{if\_} \encode{b}
        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$}
\end{align*}
The metavariables $x$ and~$n$ stand for a variable and an integer,
respectively.
This pre-encoding is just like how we represent object terms in the
metalanguage in the preceding sections, but it produces
terms in the object language rather than the metalanguage.

To evaluate~$\encode{e}$, then, we
instantiate the free variables in~$\encode{e}$ such as $\ident{int}$,
$\ident{lam}$, and $\ident{add}$ by their definitions above.  For
example, the familiar object term $(\fun{x}x)\True$ pre-encodes to
\begin{equation*}
    \encode{(\fun{x}x)\True} = \ident{app}
    (\ident{lam} (\fun{x} x))\, (\ident{bool} \True),
\end{equation*}
and to evaluate this pre-encoded term is to evaluate the object term
\begin{equation*}
    (\fun{f} \fun{x} fx) \,
    ((\fun{f} f) (\fun{x} x))\, ((\fun{b} b) \True).
\end{equation*}
Because the evaluator above mostly consists of glorified identity
functions, we expect our simple partial evaluator to reduce the
result of this instantiation to~$e$.  In general, to
interpret~$\encode{e}$ using an interpreter is to instantiate its free
variables by that interpreter's definitions.

@@

Any approach to self\hyp interpretation needs to spell out first how to
encode object terms~$e$ to object terms~$\Encode{e}$, and then how to
interpret~$\Encode{e}$ in the object language.  For our approach, we
want to define encoding in terms of pre-encoding, and interpretation
using some notion of instantiation.  Unfortunately, the simple type
structure of our object language hinders both tasks.  To continue with
the example term above, we could try to define
\begin{equation*}
    \Encode{e} =
    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}} \encode{e},
\end{equation*}
in particular
\begin{equation*}
    \Encode{(\fun{x}x)\True} =
    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool} \True).
\end{equation*}
To type-check this encoded term, we give the variable $\ident{lam}$ the
simple type $(\BB\to\BB)\to\BB\to\BB$.
We then define the self\hyp interpreter
\begin{equation*}
    \si = \fun{e} e
    (\fun{f} \fun{x} fx)
    (\fun{f} f)
    (\fun{b} b)
\end{equation*}
and apply it to the encoded term.  The result is the object term
\begin{multline*}
    \bigl(\fun{e} e (\fun{f} \fun{x} fx) (\fun{f} f) (\fun{b} b)\bigr)
\\
    \bigl(
    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool} \True)
    \bigr),
\end{multline*}
which partially evaluates to~$\True$ easily.
However, encoding fails on an term with multiple $\lambda$\hyp
abstractions at different types.  For example, the pre-encoding
\begin{equation*}
    \encode{\fun{f}\fun{x}fx}
    = \ident{lam} (\fun{f} \ident{lam} (\fun{x} \ident{app} f x))
\end{equation*}
does not type-check in any typing environment, because $\ident{lam}$ needs
to take two incompatible types.  In sum, we need more polymorphism in the
object type system to type $\ident{lam}$, $\ident{app}$, $\ident{fix}$,
and~$\ident{if\_}$ (and $\Encode{e}$ and~$\si$).
(The polytypes in |Symantics| given by Haskell's type classes and OCaml's
modules supply this polymorphism.)  Moreover, we need to encode any
polymorphism of the object language \emph{into} the object language to achieve
self\hyp interpretation.

@@

Instead of adding higher-rank and higher-kind polymorphism to our object
language (along with polymorphism over kinds!), we add let-bound polymorphism.
As usual,
we can add a new typing rule
\begin{equation*}
    \begin{prooftree}
        e_1:t_1 \quad \subst{e_2}{x}{e_1}:t_2
        \justifies \be{x=e_1} e_2 : t_2
    \end{prooftree}
    .
\end{equation*}
The pre-encoding of a let\hyp expression is trivial.
\begin{align*}
    \encode{\be{x=e_1}e_2} \quad &= \quad \be{x=\encode{e_1}} \encode{e_2}
\intertext{A \emph{context} is an object term with a hole~$[~]$.  The
hole may occur under a binder, so plugging a term into the context may
capture free variables of the term.  By pre-encoding a hole to a hole,
we extend pre-encoding from a translation on terms to one on
contexts.}
    \encode{[~]} \quad &= \quad [~]
\end{align*}
We define an interpreter in the object language to be not a term but
a context.  For example, the evaluator is the context
\begin{align*}
    \si[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} x&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y&} \\
        \be{\ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}&} \\
        \be{\ident{lam} &= \fun{f} f&} \\
        \be{\ident{app} &= \fun{f} \fun{x} fx&} \\
        \be{\ident{fix} &= \fun{g} \fix{f} \fun{x} gfx&} [~],
    \end{tabular}
\intertext{and the size\hyp measurer is the context}
    \mathrm{SZ}[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} 1&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y+1&} \\
        \be{\ident{if\_}&= b + t\,0 + e\,0&} \\
        \be{\ident{lam} &= \fun{f} f\,0 + 1&} \\
        \be{\ident{app} &= \fun{f} \fun{x} f + x + 1&} \\
        \be{\ident{fix} &= \fun{g} g\,0 + 1&} [~].
    \end{tabular}
\end{align*}
To interpret an object term~$e$ using an interpreter $I[~]$ is to
evaluate the object term~$I[\encode{e}]$.  $\si$ is
a self\hyp interpreter in the following sense.
\begin{prop}
    $\si[\encode{e}]$ is observationally equivalent to~$e$.
\end{prop}
As a corollary, we can pre-encode the self\hyp interpreter itself as
a context: the term $\si[\encode{\si[\encode{e}]}]$ is observationally
equivalent to $\si[\encode{e}]$, and in turn to~$e$.  In other words,
$\si$ can interpret itself.  Our partial evaluator is optimal with respect to
the self\hyp interpreter~$\si$.
\begin{prop}
    Let $\pe$ be the partial evaluator~|P| in~\S\ref{S:PE-solution}.
    Then the object terms $\pe(\si[\encode{e}])$ and~$\pe(e)$ are
    either both undefined or both defined and
    equal up to $\alpha$\hyp conversion.
\end{prop}

@@

We type-check a pre-encoded term~$\encode{e}$ and its interpreter~$I[~]$
only together, never separately.  This treatment has the drawback that
we must duplicate a pre-encoded term in order to interpret it in
multiple ways.  The meta-notions of contexts and plugging may seem ad
hoc, but in fact they are just reflections of the type-class and module
machinery that we have been using in this paper all along.

@@

\citet{Danvy-tagging-encoding} discuss Jones optimality at length.
There, the authors demonstrate the usefulness of HOAS for typed
self-interpretation.  However, no claim is made that the source program
is itself typed -- just that the interpreter is.  Our reading is that
the source language is not typed.  However, in sharp contrast with 
their interpreter, none of our evaluators can product pattern-match errors
(unlike the code that accompanies \citet{Danvy-tagging-encoding}).
In other words, their language (and interpreter) does use tags, however
the shift to HOAS allows the partial evaluator to remove all the tags.
In contrast, we have no tags at all.
